---
title: "Simpson's Paradox in Palmer Penguins"
format:
  html:
    toc: true
execute: 
  echo: false
---

```{r}
#| label: load
#| message: false
targets::tar_load(penguin_models_augmented)
targets::tar_load(penguin_models_summary)

library(tidyverse)
```

This is an example analysis of penguins on the Palmer Archipelago in Antarctica.

## Models

We analyzed the relationship between bill length and depth using linear models. One model did not take into account species, while the other did.

## Results

```{r}
#| label: results-stats
mod_stats <-
  penguin_models_summary %>%
    mutate(across(c(r.squared, adj.r.squared), ~round(.x, 3))) %>%
    split(.$model)
```

The model that did not account for species shows a negative relationship between bill length and depth overall (@fig-combined-plot).

The model that did account for species shows the opposite result: a positive relationship between bill length and depth for each species (@fig-separate-plot).

Furthermore, the combined model explained less of the variation in the data (adjusted *R* squared, `r mod_stats$combined$r.squared`) than the separate model (adjusted *R* squared, `r mod_stats$separate$r.squared`).

## Conclusion

The penguins data is an example of "Simpson's Paradox," whereby aggregation of data can hide trends that are happening at more nested levels. We need to be careful of this when analyzing data.

## Figures

```{r}
#| label: fig-combined-plot
#| fig-cap: "Combined model"
ggplot(
  filter(penguin_models_augmented, model == "combined"),
  aes(x = bill_length_mm)
) +
  geom_point(aes(y = bill_depth_mm)) +
  geom_line(aes(y = .fitted)) +
  labs(
    y = "Bill length (mm)",
    x = "Bill depth (mm)"
  )
```

```{r}
#| label: fig-separate-plot
#| fig-cap: "Separate model"
ggplot(
  filter(penguin_models_augmented, model == "separate"),
  aes(x = bill_length_mm, color = species)
) +
  geom_point(aes(y = bill_depth_mm)) +
  geom_line(aes(y = .fitted)) +
  labs(
    y = "Bill length (mm)",
    x = "Bill depth (mm)"
  ) +
  theme(legend.position = "bottom")
```
